\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{hyperref}
\usepackage[a4paper,margin=1cm]{geometry}
\pagestyle{empty}

\begin{document}

\hfil {\Huge Training Project Laboratory: C compiler} \hfil

\section{Scope of the project}
The goal of the project is to implement a C compiler. We will be implementing a
subset of the C11~\cite{c11} language. (This means, that any code successfully
compiled by our compiler most conform to the standard. On the other hand, we
won't necessarily be able to compile arbitrary standard conforming code. In
particular, making a self compiling compiler is out of scope for this project.)

Doing anything more than trivial optimizations are also out of scope. Making an
optimizing compiler would be a significantly harder task.

\section{Tools}
It is possible that more tools will end up being required than the ones listed
below, but we try to cover as many here as possible.
\subsection{Version control}
The GIT~\cite{git} version control system will be used for managing the entire
project. The repository will be published on the GitLab~\cite{gitlab} instance
hosted by KSZK~\cite{kszk}. The repository will be available at
\url{https://git.sch.bme.hu/gyuri/c_compiler}.
\subsection{Build system}
We will be using the Meson~\cite{meson} build system. \TeX~Live~\cite{texlive}
will be used to compile \LaTeX{} sources to PDF.
\subsection{Programming language}
The compiler will be written in the C11~\cite{c11} language.
\subsection{Lexing and parsing}
Lexer and parser generators massively simplify the processing of complex
grammars. We are choosing flex~\cite{flex} and bison~\cite{bison} because of
their very easy use with C code.
\subsection{Documentation}
We will be using \LaTeX~\cite{latex} and Markdown to write any documentation
required for the project. For markdown, we will use CommonMark~\cite{md}, as it
should work well with GitLab too.

\section{License}
The code in this project will be licensed under the GNU General Public License
v3.0~\cite{gpl_v3}.

\begin{thebibliography}{1}
\bibitem{c11} \url{http://www.open-std.org/jtc1/sc22/wg14/www/docs/n1548.pdf}
\bibitem{git} \url{https://git-scm.com}
\bibitem{kszk} \url{https://kszk.bme.hu}
\bibitem{gitlab} \url{https://gitlab.com}
\bibitem{meson} \url{https://mesonbuild.com}
\bibitem{texlive} \url{https://www.tug.org/texlive}
\bibitem{flex} \url{https://github.com/westes/flex}
\bibitem{bison} \url{https://www.gnu.org/software/bison/bison.html}
\bibitem{latex} \url{https://www.latex-project.org}
\bibitem{md} \url{https://spec.commonmark.org/current}
\bibitem{gpl_v3} \url{https://spdx.org/licenses/GPL-3.0-only.html}
\end{thebibliography}
\end{document}
